\documentclass[a4paper,11pt]{article}

\usepackage[latin1]{inputenc}

\usepackage{tikz}
\usetikzlibrary{automata,arrows,topaths,calc}


\usepackage{subcaption}

\usepackage{changepage}


\usepackage{listings}
\lstset{language=erlang}

\newcommand{\nnsection}[1]{
\section*{#1}
\addcontentsline{toc}{section}{#1}
}

\begin{document}

\begin{center}
\vspace{20pt}
\textbf{\large An exercise in communication}\\
\vspace{10pt}
\textbf{Johan Montelius}\\
\vspace{10pt}
\today{}
\end{center}

\nnsection{Introduction}

This exercise is partly about breaking up a problem in layers of
abstractions and implement each layer as a process or a set of
processes. We will model our solution as communicating finite state
machines and each machine is of course implemented as an Erlang
process.

The problem that we will look at is how to create a communication
abstraction that provides addressing and FIFO delivery over a lossy
channel. The situation will look similar to that of TCP/IP but our
problem statement and solution will be much simpler.

\section{A transport abstraction}

Assume that we want to create a {\em transport service}. The service
should provide the following functionality:

\begin{itemize}

\item identity: The service should provide addresses, we should
  be able to send and receive messages to other processes given an
  address.

\item flow control: The service should take care of flow control so
  we should not be able to send more messages than the receiver can
  consume.

\item ordered delivery: Packets should be delivered in order.

\item reliable delivery: Packets should be delivered even if the
  underlying communication channel is lossy.

\end{itemize}

This is not that hard to achieve and the solution could easily be be
described in a state diagram with 15 states and 123
transitions..... The problem we would have is that, if we want to solve
all these properties in one go, we would have so many things to keep
track of that the solution would look like a bowl of spaghetti.

Instead of solving all problems at once we divide the solution into
layers, each layer providing a service that brings us closer to the
final solution. The layers that we will implement have similarities
with the ISO communication layers or the layers in the IP stack. The
layers that we will work with are:

\begin{itemize}

\item link: This layer is given, it will send a message on a
  wire but does not know very much more. There are no guarantees that the
  messages will arrive or that messages will arrive in order.

\item network: This layer will introduce {\em network addresses} so we
  can have several nodes connected together. It uses the link layer
  straight off so we do not add ordered delivery nor guarantees of
  message delivery.

\item order: This layer will add re-transmissions of
  missing messages and keep track of ordering of messages. 
  
\item flow: This layer will and solve the flow-control problem.

\end{itemize}

The TCP protocol is similar to our order and flow layers but also
implements congestion avoidance, process addressing, stream to datagram
segmentation etc. In this implementation we only provide reliability
and flow control to make things a bit easier. We will also implement
them as separate layers to show how simple each layer can be
implemented.

\section{link layer}

The link layer is give to us and provides us with the means to send
messages to a communication layer. We will to begin with connect two
nodes together but we will then implement a hub so we can connect
several nodes together.



\begin{figure}
\centering  
\begin{tikzpicture}[>=stealth', node distance=3cm, semithick, auto]

    \node[initial, state]            (init)                              {init};
    \node[state]                     (link)    [above right of=init]     {link};
    \node[accepting, state]          (final)   [below right of=init]     {final};


    \path[->]   (init)   edge       [bend left]     node      {\{connect, Dest\}}       (link)
                         edge       [bend right]    node      {quit}          (final) 
                (link)   edge       [loop above]    node      {\{send, Msg\}}          (link)
                         edge       [loop right]    node      {\#frame\{data=Msg\}}           (link)
                         edge       [bend left]     node      {quit}          (final);
\end{tikzpicture}

\caption{The link layer.}
\label{fig:link}

\end{figure}


The link layer process is, as shown in Fig.\ref{fig:link} quite
simple. We will first start the process and give it the process
identifier of the master process. We will then send it a message that
connect the process with a destination process. The link layer is then
ready to receive the following messages:

\begin{itemize}

  \item {\tt \{send, Msg\} }: A message from the master, telling the link
    process to send the message, Msg, to the destination process. The
    message is wrapped in a {\em frame} and sent using the
    regular Erlang send primitive.

    \item {\tt \#frame\{data=Msg\}}: A  {\em frame} from another link
      process. This message should be forwarded to the master process
      using the regular Erlang send primitive. The message is sent as
      is, {\tt Msg}, without being wrapped in any special data structure.
\end{itemize}

The Erlang implementation is given in Appendix \ref{app:link}. The
definition of the frame is given in a separate module, {\tt
  frame.hrl}, since we want to make it available in other
modules. Before we continue we write a small test program and connect
two nodes together.

\begin{lstlisting}
link() ->
    S = spawn(fun() -> link_sender() end),
    R = spawn(fun() -> link_receiver() end),
    {ok, Ls} = link:start(S),
    {ok, Lr} = link:start(R),
    Ls ! {connect, Lr},
    Lr ! {connect, Ls},    
    S ! {connect, Ls},
    R ! {connect, Lr},
    ok.
\end{lstlisting}  

This program will start two processes, the {\tt link\_sender} and the
{\tt link\_receiver}, that will for example send a message to and fro.
Note the order in which we have to start the processes. The two master
processes are started first in order to provide the link processes
with their respective master. Once the link processes are created we
can connect them together and then connect the master processes to the
link processes.

The {\tt link\_sender} process could look like follows. Implement the
corresponding receive process and see if your systems runs.


\begin{lstlisting}
link_sender() ->
    receive
	{connect, N} ->
	    io:format("sender connected, sending hello~n", []),
	    N ! {send, hello},
	    receive
		Msg ->
		    io:format("sender received ~w~n", [Msg])
	    end
    end.
\end{lstlisting}

\section{a hub}

If we only have two nodes in our network it is rather pointless, so we
will implement a ``hub'' to which we can connect several nodes. The hub
will accept new connection and forward incoming frames to all
connected nodes. The state machine is as shown in Fig.~\ref{fig:hub}
simple and in the implementation we only need to solve how connection
should be stored and how to implement broadcasting of messages.

As an extra precaution we add monitoring of connected processes. If a
connected process dies we remove it from the set of connected
processes. This will save the state of the hub from growing in an
environment where nodes fail before having been disconnected.

\begin{figure}
\centering  
\begin{tikzpicture}[>=stealth', node distance=3cm, semithick, auto]

    \node[initial, state]            (hub)                              {hub};
    \node[accepting, state]          (final)   [right of=hub]     {final};


    
    \path[->]   (hub)   edge       [loop above, every loop/.append style={looseness=10,in=100, out=160 }]     node      {\{connect, Dest\}}       (hub)
                        edge       [loop below, every loop/.append style={looseness=10,in=260, out=200 }]     node      {\{disconnect, Dest\}}    (hub)
                        edge       [loop above, every loop/.append style={looseness=6,in=20, out=80 }]     node      {\#frame\{\}}                 (hub)
                        edge       [loop below, every loop/.append style={looseness=6,in=340, out=280 }]     node      {'DOWN'}                 (hub)
                        edge       []               node       {quit}          (final);
    

\end{tikzpicture}
\caption{The hub.}
\label{fig:hub}
\end{figure}


\section{network layer}

So we now have a link layer that works but the thing is that the hub
will happily broadcast any message to all connected nodes. We need a
network layer that introduces an addressing scheme so we can sen a
message to a particular node in the network.

In Fig.\ref{fig:net} we have the state diagram of the network layer
and as you see it is not much more complicated than then link
layer. You might wonder why we did not implement the network layer
directly since the link layer does not give us any thing besides
message delivery but we might want to add more features, such as error
detection, in the link layer and want to be able to do this without
modifying the network layer.

\begin{figure}
\centering  
\begin{tikzpicture}[>=stealth', node distance=3cm, semithick, auto]

    \node[initial, state]            (init)                              {init Id};
    \node[state]            (net)    [above right of=init]               {net Id};
    \node[accepting, state]          (final)   [below right of=init]     {final};

    
    \path[->]   (init)   edge       [bend left]     node      {\{connect, Lnk\}}       (net)
                         edge       [] node {quit} (final)
                (net)    edge       [loop above, every loop/.append style={looseness=10,in=20, out=80 }]     node[anchor=west]      {\#msg\{dst=Id, msg=Msg\}}     (net)
                         edge       [loop above, every loop/.append style={looseness=10,in=100, out=160 }]     node[anchor=east]      {\{send, To, Msg\}}     (net)
                         edge       [loop below, every loop/.append style={looseness=6,in=340, out=280 }]     node      {\#msg\{\}}     (net)
                         edge       []               node       {quit}          (final);
    

\end{tikzpicture}
\caption{The network layer.}
\label{fig:net}
\end{figure}

The network layer is given a unique identifier when it is started and
it will use this address to filter incoming messages. When we start the
process we also provide its master process so that it know to whom it
should redirect incoming messages. Before we start sending messages we
connect the processes to a link layer process; following the pattern of
the test procedures for the link layer.


\section{order and reliability}

In the next layer we will create a process for one particular flow of
messages. We will then make sure that these messages are delivered
properly by adding a sequence number to the messages and re-transmit
messages that could be lost. 

The big question is how we detect that a message has been lost; if
messages can come out of order and possibly be lost we need a way to
impose some order. The ordering is solved if the sending process
numbers the messages as they are sent and the receiver will buffer
received messages and deliver them in the right order. 

A problem occurs if a message is lost; if we do not do anything to
solve this the receiver will wait forever for a message that will
never arrive. The solution is of course to let the sender keep a copy
of each message in case it is lost and to let the receiver request a
re-transmission. This can be solved either by explicitly request
transmissions or by acknowledge messages that do arrive correctly, we
will choose the latter alternative.

To understand how the layer works we describe the process of a sender
process and the receiver process as two different state machines. In
the implementation we will however combine these machines into one
Erlang process.

\begin{figure}
\begin{adjustwidth}{-3cm}{-3cm}
\centering
\begin{subfigure}[b]{6cm}
\resizebox{6cm}{!}{
\begin{tikzpicture}[scale=1, >=stealth', node distance=4cm, semithick, auto]
    \node[initial, state]            (init)                              {init To};
    \node[state]                     (ready)   [above right of=init]     {ready I};
    \node[state]                     (send)   [below right of=ready]      {ack A, I};
    \node[accepting, state]          (final)   [below right of=init]     {final};

    
    \path[->]   (init)   edge       [bend left]     node      {\{connect, Net\}}       (ready)
                         edge       [bend right] node[anchor=east] {quit} (final)
                (ready)  edge       [] node[anchor=west, near start]      {\{send, Msg\}}     (send)
                         edge       [loop above]  node  {\#ack\{\}}  (ready)
                         edge       [bend right]               node[anchor=east]       {quit}          (final)                   
                (send)   edge       [loop below, every loop/.append style={looseness=6,in=340, out=30 }] node[anchor=west] {timeout}     (send)   
                         edge       [bend left] node[near start]  {\#ack\{id=A\}}  (ready)
                         edge       [bend left]               node       {quit}          (final);
\end{tikzpicture}}
\subcaption{sending process}
\end{subfigure}\qquad%
\begin{subfigure}[b]{6cm}
\resizebox{6cm}{!}{
\begin{tikzpicture}[scale=1,>=stealth', node distance=3cm, semithick, auto]
    \node[initial, state]            (init)                              {init From};
    \node[state]                     (receive)   [above right of=init]      {receive I};
    \node[accepting, state]          (final)   [below right of=init]     {final};

    
    \path[->]   (init)   edge       [bend left]     node      {\{connect, Net\}}       (receive)
                         edge       [below, bend right] node[anchor=east] {quit} (final)
                (receive)   edge    [loop above, every loop/.append style={looseness=6,in=20, out=80 }] node[anchor=west]   {\#dgr\{id=I\}}     (receive)
                         edge       [loop below, every loop/.append style={looseness=6,in=340, out=280 }] node[anchor=west] {\#dgr\{id=J\}\\ if $J<I$}     (receive)
                         edge       []               node       {quit}          (final);
\end{tikzpicture}}
\subcaption{receiving process}
\end{subfigure}

\caption{The ordering layer.}
\label{fig:order}
\end{adjustwidth}
\end{figure}

In Fig.\ref{fig:order} we see the state machines of the sender and
receiver processes. In its first incarnation we will only allow one
message to be sent at a time so the processes will be fairly simple.

Both processes are started and given a specified network address that
it is communication with. It is then given access to a network process
to use when sending messages. The network processes have addresses and
we here assume that the sender and receiver are the only processes
that are communicating using these addresses. The network processes are
thus dedicated to this communication channel.


In the simpler version the sender process can be in either of two
states: {\em ready} and {\em acknowledge}. In the ready state it waits
for a {\tt \{send, Msg\}} message from its user, it will then direct
the network layer to deliver a numbered datagram and move to the
acknowledge state. In this state it will wait for an acknowledgment
before it returns to the ready state. The receiving process will
simply wait for a datagram of the right sequence number, acknowledge
this, and wait for the next. The ordering is thus preserved by the
sequence numbers and the fact that the receiving process will only
deliver messages in order.

The model is of course complicated since datagrams can be delayed or
lost in the network layer. We therefore need to add a {\em timeout}
event in the sender process to re-send a possibly lost message. If the
message was actually lost no harm is done but if it is only delayed,
or the acknowledgment was lost, the datagram has now been
duplicated. The receiving process will never deliver duplicated
datagrams but we need to take care of these messages and discard them
if they exist. In the same way the sender process needs to be aware of
that we might have duplicate acknowledgments arriving and these should
then be discarded. Note that the acknowledgments have the
corresponding sequence number so we can detect the duplicates.

In this simple version we only allow one message to be outstanding at
any given time. Only when a message has been acknowledged will we send
the next message. This is of course a severe limitation; our
communication channel will be depending on the round trip thus being
very slow.

To solve this we extend the sender, as shown in Fig.~\ref{fig:multiple}, to allow several outstanding
messages at the same time. We extend the receive {\em acknowledge}
state so that we can send new messages while we are waiting for an
acknowledgment. We have too keep track of all sent messages and make
sure that we resend the right one so we keep all sent messages, with
their corresponding sequence number, in a buffer (an list of all
messages with the last message sent last).

The sender can now receive the following messages:

\begin{itemize}
\item {\tt \{send, Msg\}}: The process will send the message in a
  numbered datagram to the receiving process. It will store the
  message with its corresponding sequence number as the last element in
  the buffer. The sequence number is then updated.

\item {\tt \#ack\{id=A\}}: if $A$
  is equal to the sequence number of the first element in the buffer
  then this element is removed from the buffer.If we receive an
  acknowledgment and remove the last element from the buffer we move
  to the {\em ready} state.

\item {\tt \#ack\{id=A\}}: if $A$
  is less than the sequence number of the first element in the buffer,
  this is a duplicate acknowledgment and can be ignored.

\item {\tt timeout} : the process will resend the first element in the buffer.

\end{itemize}


Note that we have excluded one alternative from the description. A
sending process can send messages $14$,
$15$
and $16$,
and then receive an acknowledgment for $15$
- what should we do. We could of course remove the element from the
buffer immediately but in our description we defer the message until
$15$
is the first element in the buffer. Implicit deferral is a technique
that allows us to simplify the description but it also introduce
complications that we will discuss later.


\begin{figure}
\begin{tikzpicture}[scale=1, >=stealth', node distance=4cm, semithick, auto]
    \node[initial, state]            (init)                              {init To};
    \node[state]                     (ready)   [above right of=init]     {ready I, A};
    \node[state]                     (send)   [below right of=ready]      {ack A, I};
    \node[accepting, state]          (final)   [below right of=init]     {final};

    
    \path[->]   (init)   edge       [bend left]     node      {\{connect, Net\}}       (ready)
                         edge       [bend right] node[anchor=east] {quit} (final)
                (ready)  edge       [] node[anchor=west, near start]      {\{send, Msg\}}     (send)
                         edge       [loop above]  node  {\#ack\{\}}  (ready)
                         edge       [bend right]               node[anchor=east]       {quit}          (final)                   
                (send)   edge       [loop above, every loop/.append style={looseness=6,in=20, out=70 }] node[anchor=west]    {\{send, Msg\}} (send)
                         edge       [loop below, every loop/.append style={looseness=6,in=310, out=0 }] node[anchor=west] {timeout}     (send)   
                         edge       [loop above, every loop/.append style={looseness=6,in=230, out=170 }] node[anchor=north] {}     (send)
                         edge       [bend left] node  {\#ack\{\}}  (ready)
                         edge       [bend left]               node       {quit}          (final);
\end{tikzpicture}
\caption{The ordering process: multiple outstanding messages.}
\label{fig:multiple}
\end{figure}

In the description we have chosen to describe the sender and the
receiver as two state machines, which they are. In the implementation
however, we will implement them both as one Erlang process in the same
way as we have implemented the link layer or network layer. 


\section{flow control layer}

The next layer will be slightly different since we now will do
synchronous reads of messages. The previous layers have simply passed
an incoming message to the master process but now we require the
master process to actively do a read operations. Similarly the send
operation is acknowledged so the user of the sending process knows
that it is safe to send the next message. The application layer thus has a message interface that looks like follows:

\begin{itemize}
\item {\tt \{send, Msg, Pid\}} : sent to the sending process, a {\tt
    ok} message is delivered to the process {\tt Pid}, when its safe
  to send the next message.

\item {\tt \{read, N, Pid\}} : sent to the reading process, at most
  {\tt N} messages are returned in a message {\tt \{ok, L, Mgs\}}
  where {\tt L} is the number of messages.
\end{itemize}

This layer would be easy to implement if we only allowed one
outstanding message but we would of course like to implement a
buffered system to increase the throughput. The read process thus keeps
a buffer of incoming messages that are waiting to be read. The buffer
does of course have a maximum size and the sender must be careful not
to overflow the reader with messages. Before the sender can start to
send messages it must therefore receive a {\em synchronize} message
that informs the sender of how many more messages it can send. In the
first message the reader informs the sender of the maximum size of the
buffer and it is then the responsibility of the sender to keep track of
how many more messages it can send.


\begin{figure}
\begin{adjustwidth}{-3cm}{-3cm}

\centering
\begin{subfigure}[b]{6cm}
\resizebox{6cm}{!}{
\begin{tikzpicture}[scale=1, >=stealth', node distance=4cm, semithick, auto]
    \node[initial, state]            (init)                              {init};
    \node[state]                     (ready)   [above right of=init]     {ready};
    \node[state]                     (send)   [below right of=ready]     {send S};
    \node[accepting, state]          (final)   [below right of=init]     {final};

    
    \path[->]   (init)   edge       [bend left]     node      {\{connect, Net\}}       (ready)
                         edge       [bend right] node[anchor=east] {quit} (final)
                (ready)  edge       [bend right] node[anchor=east]      {\#syn\{\}}     (send)
                         edge       [bend right] node[anchor=east]       {quit}          (final)                   
                (send)   edge       [bend right] node[anchor=west]  {\{send, Msg, Pid\}}  (ready)
                         edge       [loop below, every loop/.append style={looseness=6,out=80, in=10 }] node[anchor=east] {}     (send)   
                         edge       [loop below, every loop/.append style={looseness=6,out=340, in=280 }] node[anchor=north] {\#syn\{\}}     (send)
                         edge       [bend left]               node       {quit}          (final);
\end{tikzpicture}}
\subcaption{sending process}
\end{subfigure}\qquad%
\begin{subfigure}[b]{6cm}
\resizebox{6cm}{!}{
\begin{tikzpicture}[scale=1,>=stealth', node distance=3cm, semithick, auto]
    \node[initial, state]            (init)                              {init Size};
    \node[state]                     (receive)   [above right of=init]      {receive S};
    \node[accepting, state]          (final)   [below right of=init]     {final};

    
    \path[->]   (init)   edge       [bend left]     node      {\{connect, Net\}}       (receive)
                         edge       [below, bend right] node[anchor=east] {quit} (final)
                (receive) edge    [loop above, every loop/.append style={looseness=6,in=20, out=80 }] node[anchor=west]   {\#msg\{data=Msg\}}     (receive)
                         edge    [loop above, every loop/.append style={looseness=6,in=300, out=0 }] node[anchor=west]   {\{read, N, Pid\}}     (receive)
                         edge       []               node       {quit}          (final);
\end{tikzpicture}}
\subcaption{receiving process}
\end{subfigure}

\caption{The flow control layer.}
\label{fig:flow}
\end{adjustwidth}
\end{figure}


In Fig~\ref{fig:flow} we sea the state diagrams of the sender and
receiver. The sender keeps track of $S$,
the free space in the receiver buffer, a value that is decremented
with each send operation. If this value goes down to $0$
it will enter the {\em ready} state were it must first see a synch
message before it can accept more send operations. 

As with the ordering layer we will implement the sender and receiver
in one Erlang process. We could however have implemented them in
separate processes but we would the of course need two processes on
either side if we want to provide a duplex channel. 



\newpage
\appendix
\section{frame.hrl} \label{app:frame}
\lstinputlisting{src/frame.hrl}
\vspace{4cm}
\section{linkl.erl} \label{app:link}
\lstinputlisting{src/link.erl}
\pagebreak
\section{hub.erl} \label{app:hub}
\lstinputlisting{src/hub.erl}
\pagebreak
\section{network.erl} \label{app:net}
\lstinputlisting{src/network.erl}
\pagebreak
\section{order.erl} \label{app:ord}
\lstinputlisting{src/order.erl}
\pagebreak 
\section{flow.erl} \label{app:flow}
\lstinputlisting{src/flow.erl}
\end{document}
