\documentclass[a4paper,11pt]{article}

\input ../include/preamble.tex

\begin{document}

\title{
    \textbf{The last assignment}\\
    \large{ID1019 Programming II}
}
\author{Johan Montelius}
\date{Spring Term 2022}
\maketitle
\defaultpagestyle


\section*{Introduction}

Morse codes were used in the days of telegraphs and manual radio
communication. It is similar in the idea to Huffman coding in that it
uses fewer symbols for frequent occurring characters and more symbols
for the infrequent ones. You task is to write an encoder and a decoder
for Morse signals, encode your name and decode two secret messages.

{\bf The solution that you present should be your original code i.e. not
  copied from any source apart from the More code tree in appendix. }

\section*{Morse codes}

There are several standards for Morse codes and we will here use a
slightly extended version since we also want to code some special
character. The Morse code uses, as you probably know, long and short
symbols to encode characters (often pronounced {\em da} and {\em di}). Since
the idea is to encode frequent characters with few symbols you might
think that is identical to Huffman codes but there is a difference. In
Morse coding we have a special signal that tells us when one character
ends and the next start. The pause between characters is necessary in
order to decode a message.

As an exampe we can look at how ``ai'' is coded. The code for 'a' is
{\em di-da}, 'i' is {\em di-di} but the code for 'l' is {\em
  di-da-di-di}.  If we just had the sequence {\em di-da-di-di} we
would not know if this was ``ai'' or ``j''; we need a third signal,
the pause, to tell the difference. A sequence
{di-da-pause-di-di-pause} is then decoded as ``ai''.

How does this change the structure of our coding tree? In a Huffman
tree we only have characters in the leafs and when we hit a leaf we
know that we have a complete code for a character. In a Morse tree we
can finish anywhere along the path to a leaf. We thus have characters
in the nodes of the tree (not all nodes but almost).

\section*{Encoding}

Your first task is to implement an encoder. The Morse codes are given
in the appendix so the only thing that you need to do is to use this
information to construct an encoding table that can then be used to
encode a message. You should use the codes given since it includes some
special characters not normally found in the Morse alphabet.

\subsection*{the encode table}

If you look in the appendix the Morse codes are given represented as a
tree where one branch represents a short signal '.' and one branch the
long signal '-'.  The character of the node is the ASCII value of a
character or \verb+ :na+ if no character has the corresponding
code. Empty branches are represented by \verb+ nil+.

\begin{minted}{elixir}
@tyespec node() :: {:node, char(), node(), node()} | :nil
\end{minted}

The More trees starts like follows; a single long signal is the code
for ASCII 116 which is a 't' and a single short is the code for 101
i.e. an 'e'.

\begin{minted}{elixir}
  {:node, :na,
    {:node, 116,  ..., ...},
    {:node, 101,  ..., ...}}
\end{minted}

The tree is not the best structure to use when you encode a message so
your first task is to transform this into a form that gives you a
$O(lg(k))$ or $O(1)$ lookup operation (where $k$ is the number of
letters in the alphabet). Describe why you have chosen the
representation that you have and its characteristics. Also present how
you construct this table given the tree in the appendix.

\subsection*{the encoder}

Once you have an encoding table you should implement an encoder that
can encode a charlist (a list of ASCII values) as a Morse code
represented as a charlist of short '.', long '-' and pause ' '
characters. These could in Elixir be written using there ASCII values
(46,45,32) or the special syntax for ASCII-values (\verb+ ?.+
\verb+ ?-+ \verb+ ?/s+). 


The encoder should have a complexity of $O(n * m)$ where $n$ is the
length of the message and $m$ is the length of the Morse codes. Since
messages might be very long you should provide an implementation that
does not use stack space proportional to the length of the
message. This means that you need to provide a tail recursive
solution. The Mores codes are so short so you could have a solution
that uses stack space proportional to the code length. 

Describe your encoder and why it meets the requirements and include
the code in your report.

Show that your encoder works by encoding you name (and if you have
non-ASCII characters in your name replace them with something close
i.e. 'è' becomes 'e' etc)

\section*{Decoding}

Your next task is to implement a decoder that can take a charlist of a
Morse coded message and return the clear text message. You will of
course first need a decoding table and the table that you constructed
for the encoder might not be your best choice.

\subsection*{the decode data structure}

The datastructure that you use should give you a lookup operation that
has an $O(m)$ complexity i.e. it should be proportional to the length
of the Morse code. This means that you can not use for example a list
of all the codes and their corresponding characters since this would
require and $O(k)$ operation to search through all the codes and an
$O(m)$ operation to see if a code matches the code that you're looking
for. The situation would improve somewhat if you choose to represent
that table as a map structure but then the lookup operation would be
$O(lg(k))$ which is good but does not take advantage of the fact that
frequent occurring characters have short codes and this is something
that you want to take advantage of (very much the way how Huffman decoding
works).

So what is your choice of decoding data structure and what does your lookup
function look like?

\subsection*{the decoder}

Once you have the lookup operation you can implement the
decoder. Since the Morse encoded messages of course could be very long
your solution should not use stack space that is proportional to the
length of the messages. Describe your decoder and why it meets the
requirements, include the implementation in your report.

Decode the secret messages below. If you cut and paste the code, make
sure that you don't have carriage-return etc in the string. The string
should only contain the dash, dot and space characters (I've included
an extra space in the end so all codes end with a pause).

\begin{verbatim}
  '.- .-.. .-.. ..-- -.-- --- ..- .-. ..-- 
    -... .- ... . ..-- .- .-. . ..-- 
    -... . .-.. --- -. --. ..-- - --- ..-- ..- ... '
\end{verbatim}

\begin{verbatim}
  '.... - - .--. ... ---... .----- .----- .-- .-- 
    .-- .-.-.- -.-- --- ..- - ..- -... . .-.-.- -.-. 
    --- -- .----- .-- .- - -.-. .... ..--.. ...- .----. 
    -.. .--.-- ..... .---- .-- ....- .-- ----. .--.-- 
    ..... --... --. .--.-- ..... ---.. -.-. .--.-- 
    ..... .---- '
\end{verbatim}


\pagebreak

\section*{The Morse codes}

\begin{minted}{elixir}
def morse() do
  {:node, :na,
    {:node, 116,
      {:node, 109,
        {:node, 111,
          {:node, :na, {:node, 48, nil, nil}, {:node, 57, nil, nil}},
          {:node, :na, nil, {:node, 56, nil, {:node, 58, nil, nil}}}},
        {:node, 103,
          {:node, 113, nil, nil},
          {:node, 122,
            {:node, :na, {:node, 44, nil, nil}, nil},
            {:node, 55, nil, nil}}}},
      {:node, 110,
        {:node, 107, {:node, 121, nil, nil}, {:node, 99, nil, nil}},
        {:node, 100,
          {:node, 120, nil, nil},
          {:node, 98, nil, {:node, 54, {:node, 45, nil, nil}, nil}}}}},
    {:node, 101,
      {:node, 97,
        {:node, 119,
          {:node, 106,
            {:node, 49, {:node, 47, nil, nil}, {:node, 61, nil, nil}},
            nil},
          {:node, 112,
            {:node, :na, {:node, 37, nil, nil}, {:node, 64, nil, nil}},
            nil}},
        {:node, 114,
          {:node, :na, nil, {:node, :na, {:node, 46, nil, nil}, nil}},
          {:node, 108, nil, nil}}},
      {:node, 105,
        {:node, 117,
          {:node, 32,
            {:node, 50, nil, nil},
            {:node, :na, nil, {:node, 63, nil, nil}}},
          {:node, 102, nil, nil}},
        {:node, 115,
          {:node, 118, {:node, 51, nil, nil}, nil},
          {:node, 104, {:node, 52, nil, nil}, {:node, 53, nil, nil}}}}}}
end
\end{minted}

\pagebreak


\end{document}
