\input{../include/preamble}

\title[ID1019 Evaluation of functions]{Evaluation - case and functions}


\author{Johan Montelius}
\institute{KTH}
\date{\semester}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}{recapitulation}

\begin{itemize}
 \pause \item defining the {\em operational semantics} of a language
 \pause \item a specification for an implementation of the language
 \pause \item a tool for us to talk about the language
\end{itemize}

\end{frame}

\begin{frame}{data structures}

  We have a small domain, consisting only of atoms and a binary compound
  structures.
  \pause
  \vspace{20pt}

  \begin{tabular}{r l l}
   {\em Atoms} & = & \{a, b, c, \ldots\} \\
   {\em Structures} & = & {\em Atoms} $\cup$ \{ \{a, b\} \textbar a $\in$ {\em Structures}  $\wedge$  b $\in$ {\em Structures} \}
  \end{tabular}

  \pause
  \vspace{20pt}
  {\em We talk about atomic and compound structures.}
\end{frame}

\begin{frame}{expressions}

\begin{grammar}
<atom> ::= :a | :b | :c | \ldots

<variable> ::= x | y | z | \ldots

<literal> ::= <atom>

<expression> ::= <literal> | <variable> |  '\{' <expression> ',' <expression> '\}'

<pattern> ::= <literal> | <variable> | '\_' | '\{' <pattern> ',' <pattern> '\}'

<match> ::=  <pattern> '=' <expression>

<sequence> ::=  <expression> | <match> ',' <sequence>
\end{grammar}

\end{frame}

\begin{frame}{an environment}

\pause\vspace{20pt}
The environment holds a set of variable substitutions (bindings):
$v/s \in \sigma$, $v$ is a variable and $s$ is a structure.


\end{frame}


\begin{frame}{evaluation of expressions}

\vspace{10pt}\pause $$\frac{a \equiv s}{E\sigma(a) \rightarrow s}$$

\vspace{10pt}\pause $$\frac{v/s \in \sigma}{E\sigma(v) \rightarrow s}$$

\vspace{10pt}\pause $$\frac{ E\sigma(e_1) \rightarrow s_1 \qquad   E\sigma(e_2) \rightarrow s_2}{E\sigma(\lbrace e_1 , e_2\rbrace) \rightarrow c(s_1, s_2)}$$

\pause\vspace{20pt}{\em Evaluation can result in $\perp$, if a variable is unbound.}

\end{frame}


\begin{frame}{pattern matching}

\begin{itemize}
  \pause \item $P\sigma(a, s) \rightarrow \sigma$  if  $a \equiv s$
  \pause \item $P\sigma(\_,s) \rightarrow \sigma$  
  \pause \item $P\sigma(v, s) \rightarrow \sigma$  if \pause $ v/s \in \sigma $
  \pause \item $P\sigma(v, s) \rightarrow \lbrace v/s \rbrace \cup \sigma$ if \pause $ v/t \not\in \sigma$
   \pause \item $P\sigma(\lbrace p_1, p_2 \rbrace, \lbrace s_1, s_2 \rbrace) \rightarrow \theta$ if
   \begin{itemize}
     \pause \item  $P\sigma(p_1, s_1) \rightarrow \sigma' \wedge P\sigma'(p_2, s_2) \rightarrow \theta$
   \end{itemize}

\end{itemize}

\pause\vspace{20pt}
Pattern matching can {\em fail}. 

\end{frame}

\begin{frame}{evaluation of sequences}

$$\frac{\sigma' = \sigma \setminus \lbrace v/t \quad | \quad v/t \in \sigma \quad \wedge \quad  v \quad {\rm in} \quad p\rbrace}{S(\sigma, p) \rightarrow \sigma'}$$

\pause\vspace{20pt}


$$\frac{   
  E\sigma(e) \rightarrow t
  \qquad \sigma' = S(\sigma, p)
  \qquad P\sigma'(p, t) \rightarrow \theta
  \qquad E\theta({\rm sequence}) \rightarrow s
}{E\sigma(p = e, {\rm sequence}) \rightarrow s}$$ 


\pause\vspace{20pt}{\em First evaluate the pattern matching expression, if it succeeds then continue with the new environment.}
\end{frame}

\begin{frame}[fragile]{case expression}

\begin{verbatim}
   case  x  do
       :a ->  :foo
       :b ->  :bar
   end
\end{verbatim}

\end{frame}

\begin{frame}{case expression}

\begin{grammar}
     <expression> ::=  <case expression> | ...  

     <case expression> ::= 'case' <expression> 'do' <clauses>  'end' 

     <clauses> ::=   <clause> | <clause> ';' <clauses>

     <clause> ::=  <pattern> '->' <sequence>
\end{grammar}
\end{frame}

\begin{frame}{evaluation of case expression}


$$\frac{E\sigma(e) \rightarrow t \qquad C\sigma(t, {\rm clauses}) \rightarrow s }{E\sigma({\tt case}\ e \ {\tt do}\ {\rm clauses} \ {\tt end}) \rightarrow s}$$
  


 \vspace{20pt}\pause $C\sigma(s, {\rm clauses})$ will select one of
 the clauses based on the patterns of the clauses and then continue the
 evaluation of the sequence of the selected clause.
\end{frame}

\begin{frame}{selection of a clause}

$$\frac{
  \sigma' = S(\sigma, p) \qquad
  P\sigma'(p, s) \rightarrow \theta \qquad
  \theta \not = {\rm fail} \qquad
  E\theta({\rm sequence}) \rightarrow s}{
C\sigma(s, p \;{\rm -\textgreater}\;    {\rm sequence} ; {\rm clauses}) \rightarrow s}$$

\pause \vspace{10pt}

$$\frac{
  \sigma' = S(\sigma, p) \qquad
  P\sigma'(p, s) \rightarrow {\rm fail} \qquad
  C\sigma(s, {\rm clauses}) \rightarrow s}{
C\sigma(s, p \;{\rm -\textgreater}\;  {\rm sequence} ; {\rm clauses}) \rightarrow s}$$


\end{frame}

\begin{frame}{example}

\begin{eval}
  \pause$E\lbrace X/\lbrace a,b\rbrace\rbrace(${\tt case X of a -> a; \{\_,Y\} -> Y end}$) \rightarrow$ 
\end{eval}

\begin{eval}
   \hspace{40pt}\pause$E\lbrace X/\lbrace a,b\rbrace\rbrace(${\tt X}$) \rightarrow \lbrace a,b\rbrace$
\end{eval}

\begin{eval}
   \pause $C\lbrace X/\lbrace a,b\rbrace\rbrace(\lbrace a,b\rbrace, ${\tt \ a -> a; \{\_,Y\} -> Y}$) \rightarrow$ 
\end{eval}
\begin{eval}
   \hspace{40pt}\pause $P\lbrace X/\lbrace a,b\rbrace\rbrace( a, \lbrace a,b\rbrace) \rightarrow \mathrm{fail}$
\end{eval}

\begin{eval}
   \pause $C\lbrace X/\lbrace a,b\rbrace\rbrace(\lbrace a,b\rbrace, ${\tt \  \{\_,Y\} -> Y}$) \rightarrow$ 
\end{eval}
\begin{eval}
   \hspace{40pt}\pause $P\lbrace X/\lbrace a,b\rbrace\rbrace(\{\_, Y\}, \lbrace a,b\rbrace) \rightarrow $ \pause $\lbrace Y/b,\  X/\lbrace a,b\rbrace\rbrace$
\end{eval}

\begin{eval}
  \pause$E\lbrace Y/b, X/\lbrace a,b\rbrace\rbrace(${\tt Y}$) \rightarrow $
\end{eval}
\begin{eval}
  \pause$b$
\end{eval}
  
\end{frame}


\begin{frame}{free variables}

Are all syntactical correct sequences also valid sequences?

\pause\vspace{20pt}

A sequence must not contain any {\em free variables}.

\pause\vspace{10pt}

A free variable in a sequence, {\tt <sequence>}, is bound by the
pattern matching expressions in the sequence {\tt <patter> =
<expression>, <sequence>} if the variable occurs in the {\tt
<pattern>}.

\pause\vspace{10pt}

A free variable in a , {\tt <sequence>}, is bound by the
pattern matching expressions in the clause {\tt <pattern> -> <sequence>} if the variable occurs in the {\tt
<pattern>}.


\end{frame}

\begin{frame}{free variables}

{\tt x = :a, \{y,z\} = \{x,:b\}, \{x,y,z\}}

\pause\vspace{20pt}

{\tt \{y,z\} = \{x,:b\}, \{x,y,z\}}


\pause\vspace{20pt}


{\tt x = \{:a,:b\}, case x do \{:a,z\} -> z end}

\end{frame}


\begin{frame}[fragile]{free variables}

\vspace{20pt}

\hspace{20pt}
\begin{verbatim}
   x = {:a,:b},
   y = case x do
         {:a, z} -> {:c, z}
       end,
   {y, z}
\end{verbatim}

\vspace{20pt}{\em This is not allowed in our language, \texttt{Z} in
  \texttt{\{Y,Z\}} is a free variable. However .... \pause this is
  allowed in Elixir and Erlang (but it should not be).}

\end{frame}


 

\begin{frame}{function definitions}

We could do without named  function definitions  but it's convenient to include them in our language. 

\vspace{20pt}

Assume we have a {\em program}, $P$, that is a set of named function definitions on the form:

  $${\rm name}(v_1, ...) {\rm do}  {\rm sequence}  {\rm end}$$

where $v_1, ...$ is a sequence of unique variables called the {\em parameters} of the function. 

\pause \vspace{20pt}

Any free variable in sequence of the definition must occur in the parameters.

\end{frame}

\begin{frame}{function application}

We extend the definition of expressions, to include function application:

\pause
\begin{grammar}
<expression> ::= <fun> '(' <arguments> ') | \ldots

<arguments> ::= <expression> | <expression> ',' <arguments>

<fun> ::=  foo | bar | zot | \ldots
\end{grammar}

\pause {\em We will later extend the definition of functions but let us keep things simple for now.}

\end{frame}


\begin{frame}{evaluation of function application}

Assume we have a program $P$ that consist of a set of named functions. 

 \begin{itemize} 
  \pause\item $E\sigma({\rm name}(e_1, \ldots)) \rightarrow E\{v_1/s_1, \ldots\}({\rm sequence} )$ 
   \begin{itemize}        
           \pause\item if $E\sigma(e_i) \rightarrow s_i$ and 
           \pause\item ${\rm name}(v_1, \ldots) \rightarrow  {\rm sequence}  \in P$ 
    \end{itemize} 
 \end{itemize} 

\pause\vspace{20pt}

{\em Note that the sequence is evaluated in an environment independent of $\sigma$.}

\end{frame}

\begin{frame}[fragile]{example}

\begin{verbatim}
def append(X,Y) do
    case x do
       :nil -> Y; 
       {h, t} -> {h, append(t, y)} 
    end
end
\end{verbatim}

\pause\vspace{20pt}
evaluate $E\lbrace\rbrace( ${\tt append(\{:a,\{:b,:nil\}\}, \{:c,:nil\})}$ )$

\end{frame}


\begin{frame}{much ado about nothing}

A lot of work for something that simple - why bother, it could not have
been done differently.
\end{frame}


\end{document}
